### 2025-05-06 Chapters/paper2.tex
changes

\begin{figure}[H]
	\centering
	\includegraphics[width=0.70\textwidth]{img/paper_2/sleep event.PNG}
	\caption{Event Mapping Plot}
	\label{fig:lstm_accuracy}
\end{figure}

@@

\begin{align}
	f_t &= \sigma(W_f x_t + U_f h_{t-1} + b_f) \quad \text{(Forget Gate)} \\
	i_t &= \sigma(W_i x_t + U_i h_{t-1} + b_i) \quad \text{(Input Gate)} \\
	\tilde{c}_t &= \tanh(W_c x_t + U_c h_{t-1} + b_c) \quad \text{(Candidate Cell State)} \\
	c_t &= f_t \odot c_{t-1} + i_t \odot \tilde{c}_t \quad \text{(Cell State Update)} \\
	o_t &= \sigma(W_o x_t + U_o h_{t-1} + b_o) \quad \text{(Output Gate)} \\
	h_t &= o_t \odot \tanh(c_t) \quad \text{(Hidden State Update)}
\end{align}

@@

\begin{align}
	\overrightarrow{h}_t &= \text{LSTM}(x_t, \overrightarrow{h}_{t-1}) \\
	\overleftarrow{h}_t &= \text{LSTM}(x_t, \overleftarrow{h}_{t+1}) \\
	h_t &= [\overrightarrow{h}_t; \overleftarrow{h}_t]
\end{align}

@@

This section presents the performance evaluation of different deep learning models, using accuracy plots and confusion matrices for visual comparison.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/lstm accuracy and loss plot.png}
	\caption{LSTM Accuracy and Loss Plot}
	\label{fig:lstm_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/LSTM_cm.png}
	\caption{LSTM Confusion Matrix}
	\label{fig:lstm_cm}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/Rnn Accuracy Plot.png}
	\caption{RNN Accuracy Plot}
	\label{fig:rnn_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/RNN_cm.png}
	\caption{RNN Confusion Matrix}
	\label{fig:rnn_cm}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/BILSTM Accuaracy Plot.png}
	\caption{BiLSTM Accuracy Plot}
	\label{fig:bilstm_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/BiLSTM_cm.png}
	\caption{BiLSTM Confusion Matrix}
	\label{fig:bilstm_cm}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/deeplearning.png}
	\caption{Neural Network Accuracy Comparison}
	\label{fig:nn_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/NN_cm.png}
	\caption{Neural Network Confusion Matrix}
	\label{fig:nn_cm}
\end{figure}

@@

In this section, we compare the performance of various models, including Neural Networks, RNN, LSTM, and Bi-LSTM, based on several evaluation metrics such as accuracy, precision, recall, and F1 score.

>>>>

\begin{figure}[H]
	\centering
	\includegraphics[width=0.60\textwidth]{img/paper_2/sleep event.PNG}
	\caption{Event Mapping Plot}
	\label{fig:lstm_accuracy}
\end{figure}

@@

\begin{align}
	f_t &= \sigma(W_f x_t + U_f h_{t-1} + b_f) \quad \text{(Forget Gate)} \\
	i_t &= \sigma(W_i x_t + U_i h_{t-1} + b_i) \quad \text{(Input Gate)} \\
	\tilde{c}t &= \tanh(W_c x_t + U_c h{t-1} + b_c) \quad \text{(Candidate Cell State)} \\
	c_t &= f_t \odot c_{t-1} + i_t \odot \tilde{c}_t \quad \text{(Cell State Update)} \\
	o_t &= \sigma(W_o x_t + U_o h_{t-1} + b_o) \quad \text{(Output Gate)} \\
	h_t &= o_t \odot \tanh(c_t) \quad \text{(Hidden State Update)}
\end{align}

@@

\begin{align}
	\overrightarrow{h}t &= \text{LSTM}(x_t, \overrightarrow{h}{t-1}) \\
	\overleftarrow{h}t &= \text{LSTM}(x_t, \overleftarrow{h}{t+1}) \\
	h_t &= [\overrightarrow{h}_t; \overleftarrow{h}_t]
\end{align}

@@

\begin{figure}[H]
	\centering
	\begin{minipage}{0.66\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/deeplearning.png}
		\caption{Neural Network Accuracy and Loss Comparison}
		\label{fig:nn_accuracy}
	\end{minipage}
	\hfill
	\begin{minipage}{0.33\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/NN_cm.png}
		\caption{Neural Network Confusion Matrix}
		\label{fig:nn_cm}
	\end{minipage}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\begin{minipage}{0.66\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/lstm accuracy and loss plot.png}
		\caption{LSTM Accuracy and Loss Plot}
		\label{fig:lstm_accuracy}
	\end{minipage}
	\hfill
	\begin{minipage}{0.32\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/LSTM_cm.png}
		\caption{LSTM Confusion Matrix}
		\label{fig:lstm_cm}
	\end{minipage}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\begin{minipage}{0.66\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/Rnn Accuracy Plot.png}
		\caption{RNN Accuracy Plot}
		\label{fig:rnn_accuracy}
	\end{minipage}
	\hfill
	\begin{minipage}{0.32\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/RNN_cm.png}
		\caption{RNN Confusion Matrix}
		\label{fig:rnn_cm}
	\end{minipage}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\begin{minipage}{0.66\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/BILSTM Accuaracy Plot.png}
		\caption{BiLSTM Accuracy Plot}
		\label{fig:bilstm_accuracy}
	\end{minipage}
	\hfill
	\begin{minipage}{0.32\textwidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper_2/BiLSTM_cm.png}
		\caption{BiLSTM Confusion Matrix}
		\label{fig:bilstm_cm}
	\end{minipage}
\end{figure}

@@

This section presents a comparative analysis of multiple machine learning models applied to the sleep stage classification problem. The evaluation metrics considered include accuracy, weighted and macro averages of precision, recall, and F1-score.

\subsection{Testing  Models Overview}

\begin{table}[H]
	\centering
	\caption{Performance Comparison of Models}
	\label{tab:metrics_summary}
	\begin{tabular}{lcccccc}
		\hline
		\textbf{Model} & \textbf{Accuracy} & \textbf{Wgt. F1} & \textbf{Macro F1} & \textbf{Macro Prec.} & \textbf{Macro Rec.} \\
		\hline
		Random Forest          & \textbf{0.843} & \textbf{0.823} & 0.641 & \textbf{0.901} & 0.558 \\
		XGBoost                & \textbf{0.854} & \textbf{0.843} & 0.644 & \textbf{0.776} & 0.599 \\
		Neural Network         & \textbf{0.777} & \textbf{0.750} & 0.487 & 0.567 & 0.452 \\
		LSTM                   & \textbf{0.800} & \textbf{0.785} & 0.461 & 0.499 & 0.436 \\
		BiLSTM                 & \textbf{0.811} & \textbf{0.803} & 0.472 & 0.494 & 0.457 \\
		RNN                    & 0.706 & 0.690 & 0.332 & 0.381 & 0.345 \\
		AdaBoost               & 0.312 & 0.353 & 0.273 & 0.323 & 0.399 \\
		KNN (with PCA)         & 0.194 & 0.189 & 0.272 & 0.296 & 0.359 \\
		\hline
	\end{tabular}
\end{table}

\subsection{Discussion}

Among all the models tested, the \textbf{XGBoost classifier} achieved the highest overall accuracy (0.854) and weighted F1-score (0.843), narrowly outperforming the Random Forest model. While Random Forest achieved a slightly higher macro precision (0.901 vs. 0.776), its recall across minority classes such as N1 and REM was relatively lower.

The deep learning-based models (LSTM and BiLSTM) followed closely behind the tree-based models, with BiLSTM slightly outperforming LSTM (macro F1 of 0.472 vs. 0.461). The \textbf{Neural Network model}, although simpler, showed moderate performance (macro F1 of 0.487), better than AdaBoost and KNN.

\textbf{K-Nearest Neighbors (KNN)} and \textbf{AdaBoost} performed poorly in this context. KNN failed to handle the class imbalance effectively, leading to very low overall accuracy (0.194). AdaBoost performed slightly better but still struggled to generalize, especially for majority class (Wake) classification.

The \textbf{RNN model} had the lowest macro F1-score (0.332) among deep models, indicating potential vanishing gradient issues or insufficient context capture compared to LSTM-based architectures.

\subsection{Conclusion}

Tree-based ensemble methods (XGBoost, Random Forest) consistently outperformed others across all key metrics. Deep learning models, especially BiLSTM, offer competitive alternatives with relatively strong performance. Simpler models like KNN and AdaBoost underperform and are not suitable for this imbalanced multiclass classification task without further tuning or class balancing.

### 2025-05-06 Chapters/paper3.tex
changes

@@

The preprocessing begins with reading the EDF recordings and assigning sleep labels based on the mapped classes. Subsequently, we applied a bandpass filter to isolate relevant frequency ranges. Specifically, the signal was filtered to retain only the 0.3–30 Hz band using the `firwin` filter design, effectively removing noise and irrelevant frequencies outside this band. Though not a standard approach, this customized filtering was found to be better suited to our specific dataset and classification task.

@@

The filtering was implemented using a windowed time-domain design via the `firwin` method. A Hamming window was employed, configured with a 0.0194 passband ripple and 53 dB stopband attenuation. The filter characteristics were as follows: lower passband edge at 0.30 Hz, lower transition bandwidth of 0.30 Hz (with a $-6$ dB cutoff at 0.15 Hz), upper passband edge at 30.00 Hz, and an upper transition bandwidth of 7.50 Hz (with a $-6$ dB cutoff at 33.75 Hz). These settings were carefully chosen to preserve the signal components relevant to sleep staging while minimizing noise.

@@

The training process for the SleepGCN-Transformer model is designed to handle class imbalance, prevent overfitting, and improve generalization. Below, we outline the training methodology using key equations:

\textbf{Label Smoothing:} To reduce overconfidence in predictions and improve generalization, we apply label smoothing:
\begin{equation}
	y_{\text{smooth}} = y(1 - \epsilon) + \frac{\epsilon}{C}
\end{equation}

\textbf{Combined Loss Function:} We incorporate both Focal Loss and smoothed labels:
\begin{equation}
	L = \alpha(1 - p)^\gamma (-y_{\text{smooth}} \log p)
\end{equation}

\textbf{Focal Loss (for class imbalance):}
\begin{equation}
	FL(p_t) = -\alpha(1 - p_t)^\gamma \log(p_t)
\end{equation}

\textbf{Class Weight Computation:} Each class receives a weight based on its frequency:
\begin{equation}
	w_c = \frac{N}{\sqrt{K + 1}}
\end{equation}

\textbf{AdamW Optimizer:} We update parameters using the AdamW optimization rule:
\begin{equation}
	\theta_{t+1} = \theta_t - \eta (\nabla\phi(\theta_t) + \lambda\theta_t)
\end{equation}

\textbf{AdamW with Decoupled Weight Decay:}
\begin{equation}
	\theta_t' = \theta_t - \eta \nabla\phi(\theta_t)
\end{equation}
\begin{equation}
	\theta_{t+1} = \theta_t' - \eta \lambda \theta_t
\end{equation}

\textbf{Learning Rate Scheduling:} Cosine Annealing is used to adjust learning rates dynamically:
\begin{equation}
	\eta_t = \eta_{\min} + \frac{1}{2} (\eta_{\max} - \eta_{\min}) \left(1 + \cos\left(\frac{\pi T_{\text{cur}}}{T_{\text{max}}}\right)\right)
\end{equation}

\textbf{Pipeline Summary:}
\begin{itemize}
	\item Load preprocessed EEG dataset for training and validation.
	\item Compute class weights.
	\item Apply label smoothing and Focal Loss.
	\item Optimize with AdamW and update learning rate via cosine scheduler.
	\item Monitor loss and accuracy metrics per epoch.
\end{itemize}

@@

\begin{center}
	\begin{minipage}[t]{0.48\textwidth}
		\begin{align}
			\text{Accuracy} &= \frac{TP + TN}{TP + TN + FP + FN} \\
			\text{Precision} &= \frac{TP}{TP + FP} \\
			\text{Recall} &= \frac{TP}{TP + FN} \\
			\text{F1\text{-}Score} &= 2 \cdot \frac{\text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}} \\
			\text{Support} &= TP + FN \\
			\text{Specificity} &= \frac{TN}{TN + FP}
		\end{align}
	\end{minipage}
	\hfill	
	\begin{minipage}[t]{0.48\textwidth}
		\begin{align}
			\text{Macro Precision} &= \frac{1}{C} \sum_{i=1}^{C} \frac{TP_i}{TP_i + FP_i} \\
			\text{Weighted Precision} &= \sum_{i=1}^{C} \frac{n_i}{N} \cdot \frac{TP_i}{TP_i + FP_i} \\
			\mathcal{L}_{CE} &= -\sum_{i=1}^{C} y_i \log(\hat{y}_i) \\
			\mathcal{L}_{FL} &= -\alpha_t (1 - p_t)^\gamma \log(p_t)
		\end{align}
	\end{minipage}
\end{center}

@@

The progression of model performance across selected epochs is summarized below:
\begin{itemize}
	\item Epoch 1: Train Loss = 0.1915, Train Accuracy = 82.91\%; Validation Loss = 0.2110, Validation Accuracy = 83.84\%
	\item Epoch 2: Train Loss = 0.1695, Train Accuracy = 87.74\%; Validation Loss = 0.2168, Validation Accuracy = 69.46\%
	\item Epoch 3: Train Loss = 0.1669, Train Accuracy = 88.15\%; Validation Loss = 0.1803, Validation Accuracy = 88.39\%
	\item Epoch 4: Train Loss = 0.1659, Train Accuracy = 88.38\%; Validation Loss = 0.1534, Validation Accuracy = 91.83\%
	\item Epoch 5: Train Loss = 0.1621, Train Accuracy = 89.45\%; Validation Loss = 0.1615, Validation Accuracy = 89.11\%
	\item Epoch 10: Train Loss = 0.1533, Train Accuracy = 91.03\%; Validation Loss = 0.1471, Validation Accuracy = 91.16\%
	\item Epoch 15: Train Loss = 0.1458, Train Accuracy = 92.59\%; Validation Loss = 0.1484, Validation Accuracy = 90.98\%
	\item Epoch 20: Train Loss = 0.1413, Train Accuracy = 93.12\%; Validation Loss = 0.1390, Validation Accuracy = 93.04\%
\end{itemize}

\begin{figure}[htbp]
	\centering
	\begin{minipage}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{img/paper_3/3D acc loss plot}
		\caption{3D Accuracy vs. Loss}
		\label{fig:3d-acc-loss-plot}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{img/paper_3/3d gradient}
		\caption{3D Gradient visualization}
		\label{fig:3d-gradient}
	\end{minipage}
\end{figure}

@@

@@

@@

@@

>>>>

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{"img/paper_3/hypnogram sleep stage transition"}
	\caption{Hypnogram With Transition Heighlights }
	\label{fig:hypnogram-sleep-stage-transition}
\end{figure}

@@

The preprocessing begins with reading the EDF recordings and assigning sleep labels based on the mapped classes. Subsequently, we applied a bandpass filter to isolate relevant frequency ranges. Specifically, the signal was filtered to retain only the 0.3–30 Hz band using the firwin filter design, effectively removing noise and irrelevant frequencies outside this band. Though not a standard approach, this customized filtering was found to be better suited to our specific dataset and classification task.

@@

The filtering was implemented using a windowed time-domain design via the firwin method. A Hamming window was employed, configured with a 0.0194 passband ripple and 53 dB stopband attenuation. The filter characteristics were as follows: lower passband edge at 0.30 Hz, lower transition bandwidth of 0.30 Hz (with a $-6$ dB cutoff at 0.15 Hz), upper passband edge at 30.00 Hz, and an upper transition bandwidth of 7.50 Hz (with a $-6$ dB cutoff at 33.75 Hz). These settings were carefully chosen to preserve the signal components relevant to sleep staging while minimizing noise.

@@

\textbf{Fully Connected Layer:}  
The final stage of the model includes a linear transformation layer that maps a 256-dimensional input vector to 5 output logits, corresponding to the five target sleep stages. This transformation enables direct classification output.

\subsubsection*{Why Focal Loss Instead of Standard Cross-Entropy?}

Standard cross-entropy loss treats all training samples equally, which can lead to a bias toward majority classes in imbalanced datasets. In such settings, the model tends to suppress the prediction of minority classes, resulting in poor generalization. Focal Loss addresses this by dynamically adjusting the contribution of each sample based on the confidence of its prediction. This mechanism reduces the influence of well-classified samples and increases the focus on harder, misclassified instances.

Focal Loss introduces two key parameters: the focusing parameter $\gamma$, which controls the down-weighting of easy examples, and the class weighting factor $\alpha$, which helps balance class frequencies during training. These features make Focal Loss particularly effective for classification tasks involving significant class imbalance.

\subsubsection*{Focal Loss Formulation}

The Focal Loss is mathematically expressed as:

\[
FL(p_t) = -\alpha(1 - p_t)^{\gamma} \log(p_t)
\]

Here, $p_t$ denotes the predicted probability for the target class, $\alpha$ is the balancing factor to compensate for class imbalance, and $\gamma$ adjusts the focus on misclassified examples.

To enhance numerical stability and improve generalization, label smoothing is applied:

\[
y_{\text{smooth}} = y(1 - \epsilon) + \frac{\epsilon}{C}
\]

This prevents issues such as $\log(0)$ and ensures a more stable optimization process. The combined implementation of Focal Loss with label smoothing can be written in a PyTorch-like style as:

\[
L = \alpha(1 - p)^{\gamma} \left( - y_{\text{smooth}} \log(p) \right)
\]

\subsubsection*{Learning Rate Scheduling}

The learning rate is a critical hyperparameter that governs the pace of model updates during training. A high learning rate may cause divergence, while a low learning rate can result in slow convergence or suboptimal local minima. Therefore, adaptive learning rate scheduling is essential for efficient training.

CosineAnnealingLR is used as the scheduling strategy, as it provides a smooth and gradual reduction in the learning rate. It allows for large updates in the early stages of training and fine-tuning in later stages, thereby helping the model generalize better. The learning rate at each epoch $t$ is computed using the following cosine annealing formula:

\[
\eta_t = \eta_{\min} + \frac{1}{2} (\eta_{\max} - \eta_{\min}) \left( 1 + \cos \left( \frac{\pi T_{\text{cur}}}{T_{\text{max}}} \right) \right)
\]

In this equation, $\eta_t$ is the learning rate at epoch $t$, $\eta_{\min}$ and $\eta_{\max}$ are the minimum and maximum learning rates respectively, $T_{\text{cur}}$ is the current epoch, and $T_{\text{max}}$ is the total number of epochs.

\subsubsection*{Training Methodology Overview}

The training pipeline is encapsulated within a dedicated class named \texttt{SleepTrainer}, which manages model training, validation, and optimization. The training process begins with the computation of class weights to address data imbalance. The model is then trained using mini-batch gradient descent, where each batch is used to compute the loss and update the model parameters accordingly. Performance is periodically evaluated on a separate validation set, and the learning rate is adjusted dynamically using the scheduler.

\subsubsection*{Hyperparameters Used}

The training setup uses a batch size of 32 and an initial learning rate of 0.0003. The optimizer of choice is AdamW, with a weight decay of $1 \times 10^{-4}$ to prevent overfitting. Training is conducted for 20 epochs, and the CosineAnnealingLR scheduler is applied to adjust the learning rate over time. This configuration ensures smooth convergence and robust generalization.

\subsubsection*{Handling Class Imbalance}

EEG sleep stage data is inherently imbalanced, with some stages appearing far more frequently than others. To address this, class weights are computed and incorporated into the loss function. The weighting formula is given by:

\[
w_c = \left( \frac{\text{Total Samples}}{\text{Class Count} + 1} \right)^{0.5}
\]

This formulation ensures that less frequent classes receive proportionally higher weights, encouraging the model to give them more attention during training. These computed weights are then integrated directly into the Focal Loss function to balance the influence of each class.

\subsubsection*{Testing Data Distribution Analysis}

Balanced testing data is crucial for an unbiased evaluation of model performance. An imbalanced test set can lead to skewed metrics that do not accurately reflect the model’s generalization capabilities. In this pipeline, the testing data is curated to ensure an even distribution across all sleep stages.

@@

\begin{center}
	\begin{minipage}[t]{0.48\textwidth}
		\begin{align}
			\text{Accuracy} &= \frac{TP + TN}{TP + TN + FP + FN} \\
			\text{Precision} &= \frac{TP}{TP + FP} \\
			\text{Recall} &= \frac{TP}{TP + FN} \\
			\text{F1\text{-}Score} &= 2 \cdot \frac{\text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}} \\
			\text{Support} &= TP + FN \\
			\text{Specificity} &= \frac{TN}{TN + FP}
		\end{align}
	\end{minipage}
	\hfill	
	\begin{minipage}[t]{0.48\textwidth}
		\begin{align}
			\text{Macro Precision} &= \frac{1}{C} \sum_{i=1}^{C} \frac{TP_i}{TP_i + FP_i} \\
			\text{Weighted Precision} &= \sum_{i=1}^{C} \frac{n_i}{N} \cdot \frac{TP_i}{TP_i + FP_i} \\
			\mathcal{L}{CE} &= -\sum{i=1}^{C} y_i \log(\hat{y}_i) \\
			\mathcal{L}_{FL} &= -\alpha_t (1 - p_t)^\gamma \log(p_t)
		\end{align}
	\end{minipage}
\end{center}

@@

The progression of model performance across selected epochs is summarized below:

\begin{figure}[htbp]
	\centering
	\begin{minipage}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{img/paper_3/3D acc loss plot}
		\caption{3D Accuracy vs. Loss}
		\label{fig:3d-acc-loss-plot}
	\end{minipage}
\end{figure}

@@

This section presents the distribution of sleep stage data, which approximates a normal distribution. The dataset is well-balanced, with an equal number of samples across all sleep stages. For testing purposes, approximately 1,050 samples were selected, ensuring that each sleep stage class contains around 210 samples. This balanced distribution supports a fair and unbiased evaluation of the model's performance.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\linewidth]{"img/paper_3/shap important"}
	\caption{Bar plot illustrating EEG channel importance based on SHAP values. Higher SHAP values indicate greater influence on the model's sleep stage predictions.}

	\label{fig:shap-important}
\end{figure}

\subsubsection*{Interpretable Sleep Stage Classification: Hypnogram Transitions and Feature Importance using LIME and SHAP}

@@

To improve the interpretability of the sleep stage classification model, we employ Explainable AI (XAI) techniques. Initially, we analyze the hypnogram transition plot, which illustrates the temporal sequence of sleep stages. This visualization helps assess whether the model's predictions align with the natural and expected progression of sleep cycles.

@@

Subsequently, we utilize LIME (Local Interpretable Model-Agnostic Explanations) and SHAP (SHapley Additive exPlanations) to investigate feature importance. These tools provide both global and local explanations of the model's behavior. Specifically, they identify the EEG channels that significantly influence the model's decisions, offering insight into which features are most critical in sleep stage classification.

@@

By leveraging these interpretability techniques, we ensure that the model's predictions are not only accurate but also aligned with known physiological sleep dynamics, thereby enhancing trust and transparency in the model's outcomes.
